A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-11-18 tex/codatadp.tex
more ideas / examples

%%% For final version as well?
\usepackage[
   a4paper,
   pdftex,
   pdftitle={A Decision Procedure for Codatatypes in SMT Solvers},
   pdfauthor={Andrew J. Reynolds and Jasmin Christian Blanchette},
   pdfkeywords={},
   pdfborder={0 0 0},
   draft=false,
   bookmarksnumbered,
   bookmarks,
   bookmarksdepth=2,
   bookmarksopenlevel=2,
   bookmarksopen]{hyperref}

@@

%\title{Mechanising the Metatheory of %(Many-Sorted)
%  First-Order Logic}
%\subtitle{Infrastructure and Application to Sort Encodings}
%\title{A Mechanised Metatheory of First-Order Logic \\
%  and Its Application to Sort Erasure}
\title{A Decision Procedure for Codatatypes in SMT Solvers}

\author {Andrew J. Reynolds\inst{1} \and Jasmin Christian Blanchette\inst{2,3}}
\authorrunning {A. J. Reynolds \and J. C. Blanchette}
\institute{
\'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
\and
Inria \& LORIA, Villers-l`es-Nancy, France \\
Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken, Germany
}

@@

@@

\def\ackname{Acknowledgement}
\ourparagraph{\ackname.}
We thank Viktor Kuncak and Tobias Nipkow for making this work possible.
The research was partially supported by the Deutsche
Forschungs\-gemein\-schaft (DFG) project
\relax{Hardening the Hammer} (grant Ni\,491\slash 14-1).
% The authors are listed in alphabetical order.
% regardless of individual contributions or seniority.

>>>>

%%% For final version as well?
\usepackage[
   a4paper,
   pdftex,
   pdftitle={A Decision Procedure for (Co)datatypes in SMT Solvers},
   pdfauthor={Andrew J. Reynolds and Jasmin Christian Blanchette},
   pdfkeywords={},
   pdfborder={0 0 0},
   draft=false,
   bookmarksnumbered,
   bookmarks,
   bookmarksdepth=2,
   bookmarksopenlevel=2,
   bookmarksopen]{hyperref}

@@

%\title{Mechanising the Metatheory of %(Many-Sorted)
%  First-Order Logic}
%\subtitle{Infrastructure and Application to Sort Encodings}
%\title{A Mechanised Metatheory of First-Order Logic \\
%  and Its Application to Sort Erasure}
\title{A Decision Procedure for (Co)datatypes in SMT Solvers}

\author {Andrew J. Reynolds\inst{1} \and Jasmin Christian Blanchette\inst{2,3}}
\authorrunning {A. J. Reynolds \and J. C. Blanchette}
\institute{
\'Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Switzerland
\and
Inria \& LORIA, Villers-l\`es-Nancy, France \\
Max-Planck-Institut f\"ur Informatik, Saarbr\"ucken, Germany
}

@@

  * briefly: datatypes and why they are useful

  * less briefly: codatatypes and why they are useful
    * in Agda, Coq, Matita -- recently also in Isabelle/HOL
    * but not in SMT-LIB (2.5?)

  * single decision procedure for datatypes and codatatypes
  * datatypes are implementation following the lines of Barrett et al.
  * codatatypes were added later, motivated by CVC4 as Isabelle backend

  * setting (Section~\ref{sec:theory-of-co-datatypes}):
    * universal formulas
    * many-sorted logic
    * mutually (co)recursive types with constructors, selectors, and
      discriminators

  * codatatypes: from a theoretical and implementational point of view, like
      datatypes but:
    * codatatypes are never empty (e.g. finite streams are rejected)
    * no acyclicity (e.g. xs = lcons(0, xs) is satisfiable)
    * instead: bisimilarity

  * motivation:
    * acyclicity and bisimilarity cannot be axiomatized, so they really belong in
      a decision procedure
    * other properties can be done more efficiently by a decision procedure

  * decision procedure is described abstractly as a calculus
    (Section~\ref{sec:the-calculus})
    * includes rewriting
    * and inferences

  * integrated with Nelson-Oppen

  * implemented in CVC4 as rewriting and a theory solver
    (\ref{sec:the-theory-solver})
    * precise strategy for applying inferences
    * about 2000 lines of code, among which 1600 are shared between datatypes
      and codatatypes

  * useful both for proving and for model finding
    * in particular, acyclicity and bisimilarity are necessary for some proofs
      (and cannot be axiomatized finitely)
    * and for model finding, without them we quickly get spurious models

%\ref{sec:examples}

  * benchmarking is often an issue -- esp. codatatypes
  * the decision procedure is evaluated on two sets of benchmarks
    (Section~\ref{sec:experimental-results}):
    * first set: hand-crafted examples
    * second set: generated from Isabelle theories using the Sledgehammer tool

\paragraph{Related Work.}

    * point to Barrett et al. for SMT datatypes
    * additional ones since then (e.g. strings?)
    * anything about codatatypes?
      * proof assistants like Agda, Coq, etc. have them
      * Dafny, CoALP
      * also a lot of theoretical research, some of which is loosely connected,
        e.g. decision procedure for corecursive functions (Henning in Nijmegen)

\section{Theory of (Co)datatypes}
\label{sec:theory-of-co-datatypes}

  * unlike datatypes, codatatypes are never empty

\section{The Calculus}
\label{sec:the-calculus}

  * SMT background

  * Rewriter
  * Actual calculus

  * Correctness

  * no "internal search"

\section{The Theory Solver}
\label{sec:the-theory-solver}

  * Strategies
  * Implementation
    * what is an SMT theory solver
    * cooperation with Nelson-Oppen / integration into SMT solver
    * "mu" stuff

  * Enumerator?
    * for model generation

%\section{Examples}
%\label{sec:examples}

\section{Experimental Results}
\label{sec:experimental-results}

  * Constructed examples
  * Isabelle benchmarks

@@

Future work:

  * more evaluation
  * recursive and corecursive functions
    (and rewriting)

\def\ackname{Acknowledgment}
\paragraph{\ackname.}

Clark Barrett and Cesare Tinelli.

Bosses: Viktor Kuncak, Tobias Nipkow, and Stephan Merz.

The research was partially supported by the Deutsche
Forschungs\-gemein\-schaft (DFG) project
\relax{Hardening the Hammer} (grant Ni\,491\slash 14-1).
